Trends and Challenges in Toronto’s Homeless Population (2018-2024)

### 2024-09-27 paper/paper.qmd
Final commit

```{r}
#| include: false
#| warning: false
#| message: false
#| 
#### Workspace setup ####
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)

```

@@

All processes are performed through the `R` language [@citeR]. Data is simulated, tested, and cleaned using the `tidyverse` [@tidyverse] packages, which also facilitate downloading data through `opendatatoronto` [@citeopendatatorontoR]. Unneeded variables are removed and renamed using `dplyr` [@dplyr] and `tidyr` [@tidyr], while graphs are plotted with `ggplot2` [@ggplot2]. The `knitr` [@knitr] package is used to generate the final PDF report.

@@

# Loading the Cleaned Dataset #
# Generated by ChatGPT
cleaned_shelter_data <- read_csv(here::here("data", "analysis_data", "analysis_data.csv"))

# Generated by ChatGPT
cleaned_shelter_data$Date <- as.Date(paste0("01-", cleaned_shelter_data$Date), format="%d-%b-%y")
cleaned_shelter_data$year <- as.numeric(format(cleaned_shelter_data$Date, "%Y"))

```

@@

# Filter to include only the "All Population" group
total_population_data <- subset(cleaned_shelter_data, Population_Group == "All Population")

# Load ggplot2 for plotting
ggplot(total_population_data, aes(x = Date, y = Homeless_Count)) +
  geom_line(color = "blue", linewidth = 1.5) +  
  labs(x = "Date",
       y = "Homeless Population Count") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

@@

# Summarize data for all years (total homeless population by age group)
# Generated by ChatGPT
total_data <- cleaned_shelter_data %>%
  summarise(Age_Under_16 = sum(Age_Under_16, na.rm = TRUE),
            Age_16_24 = sum(Age_16_24, na.rm = TRUE),
            Age_25_34 = sum(Age_25_34, na.rm = TRUE),
            Age_35_44 = sum(Age_35_44, na.rm = TRUE),
            Age_45_54 = sum(Age_45_54, na.rm = TRUE),
            Age_55_64 = sum(Age_55_64, na.rm = TRUE),
            Age_65_Over = sum(Age_65_Over, na.rm = TRUE))

# Reshape data for plotting
total_data_long <- tidyr::pivot_longer(total_data, 
                                       cols = starts_with("Age"), 
                                       names_to = "Age_Group", 
                                       values_to = "Count")

@@

# Create a pie chart with percentage labels
ggplot(total_data_long, aes(x = "", y = Count, fill = Age_Group)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")), 
            position = position_stack(vjust = 0.5), size = 4) +
  theme_void() +
  theme(legend.position = "right") +
  labs(fill = "Age Group") + 
  scale_fill_brewer(palette = "Set2")

# Summarize data by year and age group
yearly_data <- cleaned_shelter_data %>%
  group_by(year) %>%
  summarise(Age_Under_16 = sum(Age_Under_16, na.rm = TRUE),
            Age_16_24 = sum(Age_16_24, na.rm = TRUE),
            Age_25_34 = sum(Age_25_34, na.rm = TRUE),
            Age_35_44 = sum(Age_35_44, na.rm = TRUE),
            Age_45_54 = sum(Age_45_54, na.rm = TRUE),
            Age_55_64 = sum(Age_55_64, na.rm = TRUE),
            Age_65_Over = sum(Age_65_Over, na.rm = TRUE))

# Reshape the data for plotting
yearly_data_long <- tidyr::pivot_longer(yearly_data, 
                                        cols = starts_with("Age"), 
                                        names_to = "Age_Group", 
                                        values_to = "Count")

@@

# Calculate total homeless population per year and add percentage column
yearly_data_long <- yearly_data_long %>%
  group_by(year) %>%
  mutate(Total_Count = sum(Count),
         Percent = Count / Total_Count * 100)

# Create a line chart showing the percentage of each age group over time without a title
ggplot(yearly_data_long, aes(x = year, y = Percent, color = Age_Group, group = Age_Group)) +
  geom_line(size = 1.2) +  
  labs(x = "Year", y = "Percentage of Homeless Population", color = "Age Group") +  
  scale_color_brewer(palette = "Set2") +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

```

@@

# Summarize data by year for different genders
gender_data <- cleaned_shelter_data %>%
  group_by(year) %>%
  summarise(Male = sum(Male_Count, na.rm = TRUE),
            Female = sum(Female_Count, na.rm = TRUE),
            Transgender_Non_Binary = sum(Transgender_Non_Binary_Count, na.rm = TRUE))

@@

# Reshape data for plotting and update Transgender_Non_Binary to Transgender/Non-binary
total_population_long <- pivot_longer(total_population, 
                                      cols = c("Transgender_Non_Binary", "Female", "Male"), 
                                      names_to = "Gender", 
                                      values_to = "Total_Count") %>%
  mutate(Gender = recode(Gender, "Transgender_Non_Binary" = "Transgender/Non-binary"))

@@

# Create a pie chart showing the gender distribution
ggplot(total_population_long, aes(x = "", y = Total_Count, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")), 
            position = position_stack(vjust = 0.5), size = 4, fontface = "bold", color = "gray30") +  
  theme_void() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set2")

# Reshape data for plotting gender proportions over time
gender_data_long <- pivot_longer(gender_data, 
                                 cols = c("Male", "Female", "Transgender_Non_Binary"), 
                                 names_to = "Gender", 
                                 values_to = "Count") %>%
  mutate(Gender = recode(Gender, "Transgender_Non_Binary" = "Transgender/Non-binary"))

# Calculate percentages per year
gender_data_long <- gender_data_long %>%
  group_by(year) %>%
  mutate(Total_Count = sum(Count),
         Percent = Count / Total_Count * 100)

# Create a line chart to show gender distribution over time
ggplot(gender_data_long, aes(x = year, y = Percent, color = Gender, group = Gender)) +
  geom_line(size = 1.2) +  
  labs(x = "Year", y = "Percentage of Homeless Population", color = "Gender") +  
  scale_color_brewer(palette = "Set2") +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

```

@@

# Generated by ChatGPT
# Summarize data by year for Chronic and All Population
chronic_data <- cleaned_shelter_data %>%
  filter(Population_Group == 'Chronic') %>%
  group_by(year) %>%
  summarize(homeless_count = sum(Homeless_Count))

all_population_data <- cleaned_shelter_data %>%
  filter(Population_Group == 'All Population') %>%
  group_by(year) %>%
  summarize(homeless_count = sum(Homeless_Count))

# Create Non-Chronic by subtracting Chronic from All Population
non_chronic_data <- all_population_data
non_chronic_data$homeless_count <- non_chronic_data$homeless_count - chronic_data$homeless_count
non_chronic_data$group <- 'Non-Chronic'

# Add Chronic to the dataset
chronic_data$group <- 'Chronic'
combined_data <- rbind(chronic_data, non_chronic_data)  # Combine data

@@

# Plot the stacked bar chart with Chronic at the top and add percentages
ggplot(combined_data, aes(x = year, y = homeless_count, fill = group)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(data = proportions, aes(x = year, y = homeless_count_chronic, label = paste0(percentage, "%")),
            vjust = 1.5, color = "white", size = 3) +  # Add percentage labels in white color
  labs(x = "Year", y = "Homeless Population Count", fill = "Group") +
  theme_minimal() +
  scale_fill_manual(values = c("Chronic" = "blue", "Non-Chronic" = "orange")) +
  scale_x_continuous(breaks = seq(min(combined_data$year), max(combined_data$year), by = 1))

```

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
#|
#### Workspace setup ####
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(styler)
```

@@

All processes are performed through the `R` language [@citeR]. Data is simulated, tested, and cleaned using the `tidyverse` [@tidyverse] packages, which also facilitate downloading data through `opendatatoronto` [@citeopendatatorontoR] . Unneeded variables are removed and renamed using `dplyr` [@dplyr] and `tidyr` [@tidyr] packages, while graphs are plotted with `ggplot2` [@ggplot2]. Code is formatted using `styler` [@styler], and the final PDF report is generated with `knitr` [@knitr] package. The workflow and code are based on guidelines from @tellingstorieswithdata.

@@

# Loading the Cleaned Dataset #
cleaned_shelter_data <- read_csv(here::here("data", "analysis_data", "analysis_data.csv"))

# This code was developed with assistance from ChatGPT.
# Convert date strings and extract year
cleaned_shelter_data$Date <- as.Date(paste0("01-", cleaned_shelter_data$Date), format = "%d-%b-%y")
cleaned_shelter_data$year <- as.numeric(format(cleaned_shelter_data$Date, "%Y"))
```

@@

# Filter to include only the "All Population" group
# This code was developed with assistance from ChatGPT.
total_population_data <- subset(cleaned_shelter_data, Population_Group == "All Population")

# Load ggplot2 for plotting
# This code is based on Alexander (2023).
ggplot(total_population_data, aes(x = Date, y = Homeless_Count)) +
  geom_line(color = "blue", linewidth = 1.5) +
  labs(
    x = "Date",
    y = "Homeless Population Count"
  ) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

# Summarize data for all years (total homeless population by age group)
# This code was developed with assistance from ChatGPT.
total_data <- cleaned_shelter_data %>%
  summarise(
    Age_Under_16 = sum(Age_Under_16, na.rm = TRUE),
    Age_16_24 = sum(Age_16_24, na.rm = TRUE),
    Age_25_34 = sum(Age_25_34, na.rm = TRUE),
    Age_35_44 = sum(Age_35_44, na.rm = TRUE),
    Age_45_54 = sum(Age_45_54, na.rm = TRUE),
    Age_55_64 = sum(Age_55_64, na.rm = TRUE),
    Age_65_Over = sum(Age_65_Over, na.rm = TRUE)
  )

# Reshape data for plotting
total_data_long <- tidyr::pivot_longer(total_data,
  cols = starts_with("Age"),
  names_to = "Age_Group",
  values_to = "Count"
)

@@

# Create a pie chart with percentage labels
ggplot(total_data_long, aes(x = "", y = Count, fill = Age_Group)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")),
    position = position_stack(vjust = 0.5), size = 4
  ) +
  theme_void() +
  theme(legend.position = "right") +
  labs(fill = "Age Group") +
  scale_fill_brewer(palette = "Set2")

# Summarize data by year and age group
yearly_data <- cleaned_shelter_data %>%
  group_by(year) %>%
  summarise(
    Age_Under_16 = sum(Age_Under_16, na.rm = TRUE),
    Age_16_24 = sum(Age_16_24, na.rm = TRUE),
    Age_25_34 = sum(Age_25_34, na.rm = TRUE),
    Age_35_44 = sum(Age_35_44, na.rm = TRUE),
    Age_45_54 = sum(Age_45_54, na.rm = TRUE),
    Age_55_64 = sum(Age_55_64, na.rm = TRUE),
    Age_65_Over = sum(Age_65_Over, na.rm = TRUE)
  )

# Reshape the data for plotting
yearly_data_long <- tidyr::pivot_longer(yearly_data,
  cols = starts_with("Age"),
  names_to = "Age_Group",
  values_to = "Count"
)

@@

# Calculate total homeless population per year and add percentage column
yearly_data_long <- yearly_data_long %>%
  group_by(year) %>%
  mutate(
    Total_Count = sum(Count),
    Percent = Count / Total_Count * 100
  )

# Create a line chart showing the percentage of each age group over time without a title
ggplot(yearly_data_long, aes(x = year, y = Percent, color = Age_Group, group = Age_Group)) +
  geom_line(size = 1.2) +
  labs(x = "Year", y = "Percentage of Homeless Population", color = "Age Group") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
```

@@

# This code was developed with assistance from ChatGPT.
# Summarize data by year for different genders
gender_data <- cleaned_shelter_data %>%
  group_by(year) %>%
  summarise(
    Male = sum(Male_Count, na.rm = TRUE),
    Female = sum(Female_Count, na.rm = TRUE),
    Transgender_Non_Binary = sum(Transgender_Non_Binary_Count, na.rm = TRUE)
  )

@@

# Reshape data for plotting and update Transgender_Non_Binary to Transgender/Non-binary
total_population_long <- pivot_longer(total_population,
  cols = c("Transgender_Non_Binary", "Female", "Male"),
  names_to = "Gender",
  values_to = "Total_Count"
) %>%
  mutate(Gender = recode(Gender, "Transgender_Non_Binary" = "Transgender/Non-binary"))

@@

# Create a pie chart showing the gender distribution
ggplot(total_population_long, aes(x = "", y = Total_Count, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")),
    position = position_stack(vjust = 0.5), size = 4, fontface = "bold", color = "gray30"
  ) +
  theme_void() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set2")

# Reshape data for plotting gender proportions over time
gender_data_long <- pivot_longer(gender_data,
  cols = c("Male", "Female", "Transgender_Non_Binary"),
  names_to = "Gender",
  values_to = "Count"
) %>%
  mutate(Gender = recode(Gender, "Transgender_Non_Binary" = "Transgender/Non-binary"))

# Calculate percentages per year
gender_data_long <- gender_data_long %>%
  group_by(year) %>%
  mutate(
    Total_Count = sum(Count),
    Percent = Count / Total_Count * 100
  )

# Create a line chart to show gender distribution over time
ggplot(gender_data_long, aes(x = year, y = Percent, color = Gender, group = Gender)) +
  geom_line(size = 1.2) +
  labs(x = "Year", y = "Percentage of Homeless Population", color = "Gender") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
```

@@

# This code was developed with assistance from ChatGPT.
# Summarize data by year for Chronic and All Population
chronic_data <- cleaned_shelter_data %>%
  filter(Population_Group == "Chronic") %>%
  group_by(year) %>%
  summarize(homeless_count = sum(Homeless_Count))

all_population_data <- cleaned_shelter_data %>%
  filter(Population_Group == "All Population") %>%
  group_by(year) %>%
  summarize(homeless_count = sum(Homeless_Count))

# Create Non-Chronic by subtracting Chronic from All Population
non_chronic_data <- all_population_data
non_chronic_data$homeless_count <- non_chronic_data$homeless_count - chronic_data$homeless_count
non_chronic_data$group <- "Non-Chronic"

# Add Chronic to the dataset
chronic_data$group <- "Chronic"
combined_data <- rbind(chronic_data, non_chronic_data) # Combine data

@@

# Plot the stacked bar chart with Chronic at the top and add percentages
ggplot(combined_data, aes(x = year, y = homeless_count, fill = group)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(
    data = proportions, aes(x = year, y = homeless_count_chronic, label = paste0(percentage, "%")),
    vjust = 1.5, color = "white", size = 3
  ) + # Add percentage labels in white color
  labs(x = "Year", y = "Homeless Population Count", fill = "Group") +
  theme_minimal() +
  scale_fill_manual(values = c("Chronic" = "blue", "Non-Chronic" = "orange")) +
  scale_x_continuous(breaks = seq(min(combined_data$year), max(combined_data$year), by = 1))
```
